\thetitle

### 2021-03-01 writing/mutate-interesting.tex
mutators

All mutators produce mutations with type-level mistakes, but not all such
mutations necessarily represent interesting debugging scenarios. Running a
mutation may not raise a run-time error, though produce the wrong
outputs---something the rational programmer may not recognize and certainly
cannot use for a debugging session. Even if the program execution ends in an
exceptional state, it may do so too quickly; for example, the first-order export
checks of Natural may catch the error and point directly to the faulty module.

To avoid these pitfalls, an \emph{interesting debugging scenario} is
defined as a mutation that satisfies three conditions.  First, {\em
the mutation is rejected by the type checker when all missing type
annotations are restored\/}. This first criterion validates that the
debugging scenario concerns an impedance mismatch.

Second, {\em it produces a run-time exception under Erasure as-is\/}, which
ensures that it is detectable under all three semantics.  Whether a scenario
produces a run time error largely depends on the underlying semantics but
\citet{gfd-oopsla-2019} show that a run-time error under Erasure implies a run
time error under both Transient and Natural. Thus the criterion aligns
interesting debugging scenarios with the lowest denominator, Erasure. It
excludes from consideration the most complex scenarios where only Natural or
Transient can detect a mistake and Erasure is powerless.

@@

Third, {\em the exception's stack trace mentions at least three
distinct components\/}, meaning the scenario demands a sufficiently
sophisticated debugging effort: one due to the interaction between the
buggy portion of the program with the rest of the program.  In these
cases, the rational programmer must examine at least modules to locate
the source of the faulty interaction.  In technical terms, if the
stack trace mentions at least two distinct components, the bug should
be revealed by the run-time checks governing an interaction. A third
component is required because all benchmark programs include a driver
component, which is guaranteed to be in the stack trace. In essence,
this last criterion excludes a large number of trivial-to-debug
scenarios that error immediately when their mutated component is
evaluated.

@@

The definition of interesting debugging scenarios creates a powerful filter. At
a high level, the mutators produce 16,800 mutants with at least one interesting
debugging scenario each across all of benchmarks.  Broken down by benchmark, the
mutators produce at least 40 interesting scenarios for every benchmark, and
these scenarios originate from at least four different mutators per benchmark.
Thus, the mutators result in a sizable and diverse population of scenarios for
every benchmark.  Furthermore, every mutator contributes scenarios to at least
one benchmark.  Note, though, that some mutators apply only to a few benchmarks,
because they target rather specific features.  For instance, the class-focused
mutators are mainly effective in \texttt{take5}, because it makes the most
extensive use of object-oriented Racket features.

>>>>

%% -----------------------------------------------------------------------------

A type-level mutator is {\em interesting\/} (1) if, upon removal of the type
annotations from the mutant-module, it raises a run-time error, (2) whose stack
trace contains source locations from at least three modules, and (3) if it
applies to all of the to-be-evaluated semantics. The application of an
interesting mutator yields a mutant that represents a {\em debugging scenario\/}
where the rational programmer can exploit blame information.

Let us consider the three conditions in order:
\begin{enumerate}

\item By definition, an impedance mismatch is a class between type ascription of
one module to another module's exports. Hence, the application of a mutator must
create a mutant module that fails to type check. Removing the types from this
module creates an impedance mismatch between it and the modules that import its
services, directly or indirectly. This impedance mismatch is of interest if, and
only if, running the program leads to an exception---otherwise it goes
unobserved and cannot be debugged.

\item If the evaluation o mutant module immediately raises an exception because
of the mutation, there is no work for the rational programmer. Indeed, if the
stack trace contains source pointers to two modules, the scenario is still
uninteresting. Every ordinary benchmark program comes with a {\tt main} module
that acts as a driver, whose source is guaranteed to be included in the stack
trace.  Hence, the definition of interesting mutator insists on the presence of
three different modules in the stack trace. This guarantees that the debugging
scenario demands a sufficiently sophisticated effort, due to the interaction
between the buggy module with its context.  In these cases, the rational
programmer must examine at least modules to locate the source of the faulty
interaction.

\item The goal of a comparative evaluation is to give the rational programmer a
chance to debug the same scenario using different pieces of information.  In the
case of gradual typing semantics, a meta-theorem due to Greenman and
Felleisen~\cite{gf-icfp-2018} says that if a program raises an exception under
Erasure, it also errors under all other semantics.  Hence, a comparison of blame
information insists that a debugging scenario {\em raises a run-time exception
under Erasure\/}.

@@

\end{enumerate}
\mf{It seems to me that we're going out on a limb with this claim. Isn't it true
that the ``three'' is really the result of how many modules are typed/untyped in
the context of the mutant module?}
%
Although the creation of such scenarios depends on both the mutator and the
context of the mutant module, it turns out that some of the discarded mutators 
often generate uninterresting scenarios.

@@

### 2021-03-01 writing/mutate-mutators.tex
mutators

Figure~\ref{table:mutation-ops} describes the 16 mutators used for the
experiment.  For each mutator, the figure provides a short description and
an example mutation.  Most are self-explanatory.  The first twelve apply
to most gradually typed languages, including those with classes.  The last
four target distinguishing features of Typed Racket.  Specifically, {\tt
arithmetic} may replace a {\tt +} with a {\tt -} in an attempt to change
the type of the result of the arithmetic operation. In Typed Racket, {\tt
+}'s result is a {\tt Positive-Integer} when all arguments are positive
integers~\cite{stathff-padl-12}. However, the result of {\tt -} is {\tt
Integer}. In the same spirit, the {\tt boolean} mutator aims to take
advantage of Typed Racket faithfulness to Racket's ``falsiness.'' Finally,
{\tt negate-cond} and {\tt force-cond} attempt to confuse occurrence
typing.

For the design of these mutators, the starting point is the mutators of
~\citet{lksfd-popl-2020} which in turn are borrowed from   the vast
literature on mutation testing~\cite{jia2011analysis}. However, only 2
of the previous mutators transfer as is to the context of this study, 1 is
entirely ineffective and 5 need significant modifications. For the
latter  the authors relied on their decades-long experience of making type-level mistakes in Typed
Racket, some of which take a non-trivial effort to debug.  In more detail,
the relational mutator of ~\citet{lksfd-popl-2020} does not lead to
type-level errors that Typed Racket's type system can detect, so it has
been eliminated. Similarly, the original arithmetic mutator does not
reliably lead to type errors, so it has been adapted to only make operator
swaps that can affect the type of the result. For example, changing a `/`
to `*` does not affect the type of an expression, but changing `*` to `/`
may s division by zero is detectable by Typed Racket's type system. 
In addition to careful tweaking of previous mutators, the final
set of mutators comes with 9 new ones that collectively
produce a sufficiently large and diverse set of type-level mistakes. 
For example, unlike the traditional subexpression swap mutator, 
the {\tt -id} family of mutators apply in different
contexts and utilize different pieces of non-local information about the
identifiers in a program.

As a final remark, the 16 selected mutators not only produce  type-level
mistakes  reliably but also interesting debugging scenarios. Most
mutators,  including variants of the ones
~\citet{w:good-types-mutation} propose, some times achieve the first but 
rarely the second. 
 For example,  adding a new public method to
a class surprisingly leads to mutants that either terminate 
successfully or to mutants that fail 
 immediately during the evaluation of the mutated class.  
 In contrast, the seemingly similar mutator that turns a private method into    
 a public one affects inheritance and leads to subtle bugs.  
Hence, due to the elusive nature of mutators, it is imperative to define
precisely what makes a debugging scenario interesting.

>>>>

Figure~\ref{table:mutation-ops} describes 16 mutators that satisfy this
requirement.  For each mutator, the figure provides a short description and an
example. The starting point is the mutator set of~\citet{lksfd-popl-2020}
which in turn are borrowed from the vast literature on mutation
testing~\cite{jia2011analysis}. But, only 2 of the previous mutators transfer
as is to the context of this study, 1 is entirely ineffective and 5 need
significant modifications. For the latter the authors relied on their
decades-long experience of making type-level mistakes in Typed Racket, some of
which take a non-trivial effort to debug.  In more detail, the relational
mutator of ~\citet{lksfd-popl-2020} does not lead to type-level errors that
Typed Racket's type system can detect, so it has been eliminated. Similarly,
the original arithmetic mutator does not reliably lead to type errors, so it
has been adapted to only make operator swaps that can affect the type of the
result. For example, changing a `/` to `*` does not affect the type of an
expression, but changing `*` to `/` may s division by zero is detectable by
Typed Racket's type system.  In addition to careful tweaking of previous
mutators, the final set of mutators comes with 9 new ones that collectively
produce a sufficiently large and diverse set of type-level mistakes.  For
example, unlike the traditional subexpression swap mutator, the {\tt -id}
family of mutators apply in different contexts and utilize different pieces of
non-local information about the identifiers in a program.

Most of the mutators are self-explanatory.  The first six apply to all
gradually typed languages; the next six to those that include classes and
objects.  The last four target distinguishing features of Typed Racket,
specifically is sophisticated type system. In particular, {\tt arithmetic}
replaces a {\tt +} with a {\tt -} in an attempt to change the type of the
arithmetic expression; {\tt +}'s result is a {\tt Positive-Integer} when all
arguments are positive integers, while {\tt -} yields {\tt
Integer}~\cite{stathff-padl-12}. Similarly, the other three also aim to usurp
Typed Racket's occurrence type system.
